Optimizing Abstract Abstract Machines

### 2012-11-13 paper/paper.tex
Made explicit that laziness does not change the concrete semantics, took a stab at recategorizing the techniques.

\documentclass[preprint,onecolumn,9pt]{sigplanconf} %{onecol}
\usepackage{alltt,mathpartir}
\usepackage{amsmath,amsthm}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{url}
\usepackage{graphicx}
\usepackage{balance}
\usepackage{calc}

@@

\begin{figure}
\hrule
\vspace{2mm}
\begin{alltt}
(define (zero? n) ((n (\(\lambda\) (zx) #f)) #t))
(define (((mult m1) m2) mf) (m2 (m1 mf)))
(define ((((plus p1) p2) pf) x)
  ((p1 pf) ((p2 pf) x)))

@@

\begin{figure}
\hrule
\vspace{2mm}
\begin{alltt}
(define (id x) x)
(define (f n)
  (cond [(<= n 1)  1]
        [else      (* n (f (- n 1)))]))
(define (g n)
  (cond [(<= n 1)  1]
        [else      (+ (* n n) (g (- n 1)))]))
(print (+ ((id f) 3) ((id g) 4)))
\end{alltt}
\hrule
\begin{center}
\begin{tabular}{ccc}
\raisebox{1ex-\height}{
\includegraphics[height=4in]{introspective-base.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3.7in]{introspective-lazy.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=2.2in]{introspective-lazyc.pdf}}
\\
(a) Baseline
&
(b) Lazy
&
(c) Compiled
\end{tabular}
\end{center}
\caption{Example state graphs for the program above.  Part (a) shows
  the result of the baseline analyzer.  It has long ``corridor''
  transitions and ``diamond'' subgraphs that fan-out from
  nondeterminism and fan-in from joins.  Part (b) shows the result of
  performing nondeterminism lazily and thus avoids many of the diamond
  subgraphs.  Part (c) shows the result of abstract compilation that
  removes intepretive overhead in the form of intermediate states,
  thus minimizing the corridor transitions.  The end result is a more
  compact abstraction of the program that can be generated faster.}
\end{figure}

\newpage
\
\newpage
\section{An abstract$^{\text 2}$ machine for ISWIM}
\label{sec:aam}

@@

The uniform $k$-CFA allocation strategy would make $eval$ in figure
\ref{fig:aam} a computable abstraction of reachable states, but not an
effective one. This is not the strategy that AAM, nor we,
recommend. We explain a succession of approximations to reach the
baseline analysis that we compare performance against. But first, we
need a fixpoint combinator to drive the succession of reduction
relations we define.

\subsection{Generic fixpoint calculator}
\label{sec:fixpoint}

We compute the semantics of a program by iterating the state
transition relation until a fixed point in the reachable states is
reached. We begin with Kleene's fixpoint combinator, ${\mathcal F}$,
presented in the frontier/accumulator style:

@@

The $\machstep$ relation is most naturally written as a function from
a state to a set of all next states. To calculate $eval$, we find a fixed point of
$\machstep$ lifted to step a set of states. The lifting is a natural one:

@@

\begin{align*}
(cs, \msto) &\mathrel{\widehat{\machstep}} (cs', \msto') \\
\mbox{where } cs' &= \{c' \mid wn(c, \msto) \machstep wn(c, \msto^c), c \in cs\} \\
              \msto' &= \bigsqcup\{\msto^c \mid wn(c,\msto)\machstep wn(c', \msto^c), c \in cs\} \\
wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont}
\end{align*}
That is, the next set of states is what we get by stepping each
given state, without its store. The next store is the combination of all produced
stores. For simplicity, assume that stuck states step to themselves.

@@

This extra store-allocation is effectively naming all intermediate
results, and thus the precision aligns with an analysis specialized to
ANF. We can also play a representation trick in 0-CFA and remove
$\menv$. In monovariant analyses, variables map to themselves, meaning
$\menv$ is effectively the identity function. That degeneration, in
turn, allows us to simply discard it from the analysis.

@@

We discuss a succession of techniques for increasing performance by
decreasing the interpretive overhead of abstract interpretation. They
come in two flavors: decreasing the size of the represented state
space [SS], and decreasing the amount of intermediate data structures [DS].

\begin{itemize}
 \item{[SS] lazy non-determinism: quotient by sets of values}
 \item{[SS] abstract compilation: eliminate {\tt ev} states}
 \item{[DS] store deltas: operate on store changes rather than whole stores}
 \item{[DS] timestamp: represent the store as a global, mutable hash and
   determine a state's store by the number of changes made to the
   global store at the time of visiting}
 \item{[DS] preallocation: represent the global store as a global, mutable vector}
 \item{[SS] uniform literal approximation: coarsely abstract literal data}
\end{itemize}

Some techniques preserve the precision of the underlying analysis, and
others do not. We will discuss the design decisions behind the
precision-lossy techniques.  Each technique requires minimal changes
to the reduction relation/fixpoint, and each improves performance by
integer factors.

@@

A shortcoming of the AAM approach is that much of the non-determinism
is spurious. For example, in a function application, {\tt (f x y)},
where all variables have been joined with multiple values, each
variable reference transitions to several new states that comprise the
cartesian product of all the values of {\tt f}, {\tt x} and {\tt y} (or in the our baseline analysis, the sum),
only to be joined back together in their respective application positions. We would
see transitions of the following shape:
\begin{center}
\includegraphics[scale=0.3]{fanout.pdf}
\end{center}
For this reason, we {\it delay} the non-determinism
until it is needed in {\it strict contexts} (such as the guard for an
{\tt if} a called procedure, or a numerical primitive application). We instead get transitions of this shape:
\begin{center}
\includegraphics[scale=0.3]{lazy.pdf}
\end{center}

For this, we introduce a new kind of value, $\saddr{\maddr}$, for delaying looking up $\maddr$. This acts to quotient the state space by the values at the delayed address. The following rules highlight the changes to the semantics:

@@

\paragraph{Lose precision; simplify implementation}
This semantics introduces a subtle precision difference over the
baseline. Consider a configuration where a reference to a variable and
a binding of a variable will happen in one step. With laziness, the
reference will mean the original binding(s) of the variable or the new
one, because the actual store lookup is delayed precisely one step
(i.e. laziness is administrative). Without laziness, the reference
will fan out to all the bindings of the variable before the new
binding happens and thus might have an observable precision
difference.

\paragraph{Regain precision; complicate implementation}
The administrative nature of laziness means that we could remove the
loss in precision by essentially duplicating the reduction relation to
specialize variable lookup. This works since in the semantics of ISWIM
with store-allocated results consumes the value component of states in
one step. This is not the case for semantics that replicate the value
component across reductions, say for popping off exception handler
frames. Further convolution is needed to remove the administrative
nature of laziness in these semantics. Due to the increase of
conceptual complexity for negligable benefit, we decided against this
approach.

@@

The laziness optimization saved time by doing the same amount of reasoning as before but in fewer transitions. We can exploit the same idea - same reasoning, fewer transitions - with abstract compilation. Abstract compilation transforms complex expression whose \emph{abstract} evaluation is deterministic into ``abstract bytecodes.''  The abstract interpreter then does in one transition what previously took many. Abstract compilation eliminates unnecessary allocation, deallocation and branching.

@@

The compilation step converts expressions into functions that expect
the other components of the {\tt ev} state. Its definition in figure
\ref{fig:compile} shows close similarity to the rules for interpreting
    {\tt ev} states. The next step is to change creation of {\tt ev}
    states to simply calling these functions. Figure \ref{fig:caam}
    shows the modified reduction relation.

@@

Every step the analysis makes for the above techniques requires
joining several, commonly large, stores together. Each step is
unlikely to always modify all addresses of the store, so this is
needlessly memory- and time-intensive. We can instead log store
changes and replay the change log on the full store after all steps
have completed. This uses far fewer join operations, leading to less
overhead.

We represent changes as $\msdiff \in Store\msdiff = List[Addr \times
  \Set(Storeable)]$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$ now
becomes $\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$
begins empty ($\mtlst$) for each step. Applying the changes to the
full store is straightforward - we interpret $\msdiff$ as a $Store \to
Store$ function:

\begin{align*}
\mtlst(\msto) &= \msto \\
\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}(\msto) &= \msdiff(\msto)\sqcup[\maddr \mapsto \mval{s}]
\end{align*}

@@

\begin{align*}
(cs, \msto) &\mathrel{\widehat{\machstep}} (cs\cup cs', \msdiff(\msto)) \\
\mbox{ where } (cs', \msdiff) &= step^*(\varnothing, cs, \mtlst) \\
step^*(S, \varnothing, \msdiff) &= (S, \msdiff) \\
step^*(S, \{c\}\cup cs, \msdiff &= step^*(S\cup cs^*, cs, \msdiff^*) \\
cs^* &= \{c' \mid (c,\msto,\msdiff) \machstep (c',\msdiff^c) \} \\
\msdiff^* &= concat\{\msdiff^c \mid (c, \msto,\msdiff) \machstep (c', \msdiff^c)\}
\end{align*}

@@

The new fixpoint calculation is defined in
figure~\ref{fig:imperative}.  To ensure termination, we guard adding
$c$ to {\tt todo} by the following check:
\begin{align*}\msdiff\mbox{{\tt ?}} \vee {\tt seen}(c) \neq {\tt T}\end{align*}
If it succeeds, {\tt todo} gets $c$ and we update {\tt seen} to map $c$ to {\tt T}.

After all steps complete, we apply $\msdiff$ to $\msto$ imperatively and increase
{\tt T} as long as there was at least one change in $\msdiff$. This
logic leads to termination if we know that each $\ttuple{\maddr}{\mval{s}}$
in $\msdiff$ would change the value of $\maddr$ in the current
store. Thus, we also guard additions to $\msdiff$ so that only updates
that would change the store are permitted. Each time $\msdiff$ is
successfully extended, we set {\tt $\msdiff$?} to true. Before each
individual step, we set it to false.

\begin{figure}
\begin{alltt}
eval(\(e\)) :=
  \(\msto\), todo, seen, T := \(\varnothing\), \(\varnothing\), [], 0
  \(\compile{e}\)(\(\mtcntr, \varnothing, \varnothing, \kmt\))
  while(true):
    if todo = \(\varnothing\): return (keys(seen), \(\msto\))
    else:
      let old := todo
      todo, \(\msdiff\) := \(\varnothing\), \(\mtlst\)
      foreach c \(\in\) old: \(\msdiff\)? := false; c()
      unless \(\msdiff\) = \(\mtlst\): T += 1; \(\msdiff(\msto)\)
\end{alltt}
\caption{Imperative fixpoint algorithm}
\label{fig:imperative}
\end{figure}

@@

This technique by itself only works in 0-CFA, since there are many more
addresses that can be allocated in 1-CFA and higher. To support
general $k$-CFA in a vector representation, we use a growable vector
and a map from addresses to assigned indices. When assignable indices
run out, we grow the vector. The cost of this map is less than a hash
representation for the whole store since typically there are more
references to an address than allocations of it. This is more
space-efficient than preallocating $(\mlab_{max}+1)^k$ addresses
up-front and addressing by the contour converted to polynomial, since
many combinations of labels are not likely to occur. The all up-front
allocation again breaks down when we add compound data, which
introduce new kinds of labels (e.g. left and right components of a
tuple at the contour of creation).

@@

Practical issues of performance arise when one chooses the
``obvious,'' or ``natural'' abstraction for compound data. Our running
example will be the classic Boyer benchmark. The detail of note for
the Boyer benchmark is that it has large list literals, the exact
values of which do not particularly matter for many analyses. If we
interpret a list literal of 30 elements (that again have 10 elements
each) as 300 applications of {\tt cons}, and then interpret a function
that traverses the tree, we will end up interpreting that function at
least 300 times, when once or twice would do.

@@

The number of syntactic uses of {\tt cons} versus implicit uses via
literal lists is dramatically smaller in typical Scheme programs. We
use the above abstraction for these syntactic uses, but choose to
interpret literal lists as not always sugar for cascading {\tt
  cons}es. In particular, if a list literal is ``too big'' (in our
case length $>$ 1), we interpret the list as a circular data
structure; the right address points back to the {\tt cons} itself, and
the left address points to all of the elements of the list. We take
this farther and join the list elements together in with a coarse
type-based abstraction. In effect, large lists or vectors of literal
numbers become unbounded lists/vectors of ``number.'' Heterogeneous
data is combined to just ``data,'' rather than a union of ``number'',
``string'', etc. since primitives are interpreted on every combination
of values that flow to them, and unions lead to more primitive
interpretations.

@@

Many benchmarks cause the baseline analyzer to take longer than 30
minutes or to consume more 1 gigabit of memory, at which point the
analysis is stopped.  This is the case for the largest benchmark
program, which is 3,500 lines of code and takes under a minute in the
most optimized analyzer.  For those benchmarks that did complete on
the baseline, the optimized analyzer outperformed the baseline by a
factor of 475 to 4,382.

>>>>

\documentclass[preprint,onecolumn,9pt]{sigplanconf} %{onecol}
\usepackage{alltt,mathpartir}
\usepackage{amsmath,amsthm}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{url}
\usepackage{graphicx}
\usepackage{balance}

@@

\begin{figure}
\begin{alltt}
(define (zero? n) ((n (\(\lambda\) (zx) #f)) #t))
(define (((mult m1) m2) mf) (m2 (m1 mf)))
(define ((((plus p1) p2) pf) x)
  ((p1 pf) ((p2 pf) x)))

@@

\section{An abstract$^{\text 2}$ machine for ISWIM}
\label{sec:aam}

@@

The uniform $k$-CFA allocation strategy would make $eval$ in figure
\ref{fig:aam} a computable abstraction of reachable states, but not an
effective one. This is not the strategy that AAM, nor we,
recommend. We explain a succession of approximations to reach the
baseline analysis that we compare performance against. But first, we
need a fixed-point combinator to drive the succession of reduction
relations we define.

\subsection{Generic fixed-point calculator}
\label{sec:fixpoint}

We compute the semantics of a program by iterating the state
transition relation until a fixed-point in the reachable states is
reached. We begin with Kleene's fixed-point combinator, ${\mathcal F}$,
operationalized in the frontier/accumulator style:

@@

The $\machstep$ relation is most naturally written as a function from
a state to a set of all next states. To calculate $eval$, we find a fixed-point of
$\machstep$ lifted to step a set of states. The lifting is a natural one:

@@

\begin{align*}
(cs, \msto) &\mathrel{\widehat{\machstep}} (cs', \msto') \\
\mbox{where } cs' &= \{c' \mid wn(c, \msto) \mathrel{\machstep} wn(c, \msto^c), c \in cs\} \\
              \msto' &= \bigsqcup\{\msto^c \mid wn(c,\msto)\mathrel{\machstep} wn(c', \msto^c), c \in cs\} \\
wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
That is, the next set of states is what we get by stepping each
given state, without its store. The next store is the combination of all produced
stores. For simplicity, assume that stuck states step to themselves.

@@

This extra store-allocation is effectively naming all intermediate
results, and thus the precision aligns with an analysis specialized to
ANF. We can also play a representation trick in 0-CFA and remove
$\menv$. In monovariant analyses, variables map to themselves, meaning
$\menv$ is effectively the identity function. That degeneration, in
turn, allows us to discard it from the analysis.

@@

We discuss a succession of techniques for increasing performance by
decreasing the interpretive overhead of abstract interpretation. They
come in two flavors: decreasing the size of the state space with domain-specific knowledge [DSK], and
decreasing the amount of intermediate data structures [Eng](ineering).

\begin{enumerate}
 \item{[DSK] lazy non-determinism: quotient by sets of values}
 \item{[DSK] abstract compilation: eliminate {\tt ev} states}
 \item{[Eng] store deltas: operate on store changes rather than whole stores}
 \item{[DSK/Eng] timestamp: represent the store as a global, mutable
   hash and determine a state's store by the number of changes made to
   the global store at the time of visiting}
 \item{[Eng] preallocation: represent the global store as a global, mutable vector}
 \item{[DSK] uniform literal approximation: coarsely abstract literal data}
\end{enumerate}

Some techniques preserve the precision of the underlying analysis, and
others do not. We will discuss the design decisions behind the
precision-lossy techniques.  Each technique requires minimal changes
to the reduction relation, and each improves performance by
integer factors.

@@

A shortcoming of the AAM approach is that much of the non-determinism
is spurious. For example, in a function application, {\tt (f x y)},
where all variables have been joined with multiple values, each
variable reference transitions to as many new states as there are
values, only to be joined back together in their respective
application positions. We would see transitions of the following
shape:
\begin{center}
\includegraphics[scale=0.3]{fanout.pdf}
\end{center}
For this reason, we {\it delay} the non-determinism
until it is needed in {\it strict contexts} (such as the guard for an
{\tt if} a called procedure, or a numerical primitive application). We instead get transitions of this shape:
\begin{center}
\includegraphics[scale=0.3]{lazy.pdf}
\end{center}

This does not change the concrete semantics of the language. It is an
abstraction of the otherwise several states the original
non-deterministic semantics steps to. We say the abstraction is
\emph{lazy} because it delays dereferencing an address until its
contents are \emph{needed} as values in the semantics. It does not
change the execution order that leads to the values that are stored in
the address.

We introduce a new kind of value, $\saddr{\maddr}$, that represents a
delayed lookup of $\maddr$. This acts to quotient the state space by
the values at the delayed address. The following rules highlight the
changes to the semantics:

@@

\paragraph{Lose precision; simplify implementation}
This semantics introduces a subtle precision difference over the
baseline. Consider a configuration where a reference to a variable and
a binding of a variable will happen in one step. With laziness, the
reference will mean the original binding(s) of the variable or the new
one, because the actual store lookup is delayed one step
(i.e. laziness is administrative). Without laziness, the reference
will fan out to all the bindings of the variable before the new
binding happens and thus might have an observable precision
difference.

\paragraph{Regain precision; complicate implementation}
The administrative nature of laziness means that we could remove the
loss in precision by duplicating the reduction relation to specialize
variable lookup. This works since in the semantics of ISWIM with
store-allocated results consumes the value component of states in one
step. This is not the case for semantics that replicate the value
component across reductions, say for popping off exception handler
frames. Further convolution is needed to remove the administrative
nature of laziness in these semantics. Due to the increase of
conceptual complexity for negligable benefit, we decided against this
approach.

@@

The laziness optimization saved time by doing the same amount of
reasoning as before but in fewer transitions. We can exploit the same
idea - same reasoning, fewer transitions - with abstract
compilation. Abstract compilation is \emph{precision-preserving} and
transforms complex expressions whose \emph{abstract} evaluation is
deterministic into ``abstract bytecodes.''  The abstract interpreter
then does in one transition what previously took many. Abstract
compilation eliminates unnecessary allocation, deallocation and
branching.

@@

The compilation step converts expressions into functions that expect
the other components of the {\tt ev} state. Its definition in figure
\ref{fig:compile} shows close similarity to the rules for interpreting
    {\tt ev} states. The next step is to change reduction rules that
    create {\tt ev} states to instead call these functions. Figure
    \ref{fig:caam} shows the modified reduction relation.

@@

Every step the analysis makes for the above techniques requires
joining large stores together. Not every step will modify all
addresses of the store, so joining entire stores is wasteful in terms
of memory and time. We can instead log store changes and replay the
change log on the full store after all steps have completed. This uses
far fewer join operations, leading to less overhead, and is \emph{precision preserving}.

We represent changes as $\msdiff \in Store\msdiff = List[Addr \times
  \Set(Storeable)]$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$
instead becomes a log addition
$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins
empty ($\mtlst$) for each step. Applying the changes to the full store
is straightforward:

\begin{align*}
replay &: Store\msdiff \times Store \to Store \\
replay(\mtlst, \msto) &= \msto \\
replay(\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}, \msto) &= replay(\msdiff, \msto\sqcup[\maddr \mapsto \mval{s}])
\end{align*}

@@

\begin{align*}
(cs, \msto) &\mathrel{\widehat{\machstep}} (cs\cup cs', replay(\msdiff,\msto)) \\
\mbox{ where } (cs', \msdiff) &= step^*(\varnothing, cs, \mtlst) \\
step^*(S, \varnothing, \msdiff) &= (S, \msdiff) \\
step^*(S, \{c\}\cup cs, \msdiff &= step^*(S\cup cs^*, cs, \msdiff^*) \\
cs^* &= \{c' \mid (c,\msto,\msdiff) \machstep (c',\msdiff^c) \} \\
\msdiff^* &= concat\{\msdiff^c \mid (c, \msto,\msdiff) \machstep (c', \msdiff^c)\}
\end{align*}

@@

The new $eval$ calculation is defined in
figure~\ref{fig:imperative}.  To ensure termination, we guard adding
$c$ to {\tt todo} by the following check:
\begin{align*}\msdiff\mbox{{\tt ?}} \vee {\tt seen}(c) \neq {\tt T}\end{align*}
If it succeeds, {\tt todo} gets $c$ and we update {\tt seen} to map $c$ to {\tt T}.

After all steps complete, we apply $\msdiff$ to $\msto$ imperatively
(with {\tt replay!}) and increase {\tt T} as long as there was at
least one change in $\msdiff$. This logic leads to termination if we
know that each $\ttuple{\maddr}{\mval{s}}$ in $\msdiff$ would change
the value of $\maddr$ in the current store. Thus, we also guard
additions to $\msdiff$ so that only updates that would change the
store are permitted. Each time $\msdiff$ is successfully extended, we
set {\tt $\msdiff$?} to true. Before each individual step, we set it
to false.

\begin{figure}
\begin{alltt}
eval(\(e\)) :=
  \(\msto\), todo, seen, T := \(\varnothing\), \(\varnothing\), [], 0
  \(\compile{e}\)(\(\mtcntr, \varnothing, \varnothing, \kmt\))
  while(true):
    if todo = \(\varnothing\): return (keys(seen), \(\msto\))
    else:
      let old := todo
      todo, \(\msdiff\) := \(\varnothing\), \(\mtlst\)
      foreach c \(\in\) old: \(\msdiff\)? := false; c()
      unless \(\msdiff\) = \(\mtlst\): T += 1; replay!(\(\msdiff, \msto\))
\end{alltt}
\caption{Imperative algorithm}
\label{fig:imperative}
\end{figure}

@@

This technique by itself only works in 0-CFA, since there are many
more addresses that can be allocated in 1-CFA and higher. To support
general $k$-CFA in a vector representation, we use a growable vector
and a map from addresses to assigned indices. When assignable indices
run out, we grow the vector. The cost of this map is less than a hash
representation for the whole store since typically there are more
references to an address than allocations of it. This is more
space-efficient than preallocating $(\mlab_{max}+1)^k$ addresses
up-front and addressing by the contour converted to polynomial, since
a majority of combinations of labels will not occur, except in
pathological programs. The all up-front allocation again breaks down
when we add compound data, which introduce new kinds of labels
(e.g. left and right components of a tuple at the contour of
creation).

@@

Practical issues of performance arise when one chooses the
``obvious,'' or ``natural'' abstraction for compound data. Our running
example will be the classic Boyer benchmark. The detail of note for
the Boyer benchmark is that it has large list literals, the exact
values of which do not matter for many analyses. If we
interpret a list literal of 30 elements (that again have 10 elements
each) as 300 applications of {\tt cons}, and then interpret a function
that traverses the tree, we will end up interpreting that function at
least 300 times, when once or twice would do.

@@

The number of syntactic uses of {\tt cons} versus implicit uses via
literal lists is smaller in typical Scheme programs. We use the above
abstraction for these syntactic uses, but choose to interpret literal
lists as not always sugar for cascading {\tt cons}es. In particular,
if a list literal is ``too big'' (in our case length $>$ 1), we
interpret the list as a circular data structure; the right address
points back to the {\tt cons} itself, and the left address points to
all of the elements of the list. We take this farther and join the
list elements together in with a coarse type-based abstraction. In
effect, large lists or vectors of literal numbers become unbounded
lists/vectors of ``number.'' Heterogeneous data is combined to just
``data,'' rather than a union of ``number'', ``string'', etc. since
primitives are interpreted on every combination of values that flow to
them, and unions lead to more primitive interpretations.

@@

Many benchmarks cause the baseline analyzer to take longer than 30
minutes or to consume more 1 gigabyte of memory, at which point the
analysis is stopped.  This is the case for the largest benchmark
program, which is 3,500 lines of code and takes under a minute in the
most optimized analyzer.  For those benchmarks that did complete on
the baseline, the optimized analyzer outperformed the baseline by a
factor of 475 to 4,382.
